Reviving the Lost Art of \\Combinator Graph Reduction

### 2023-02-01 experiments/superg/doc/paper/paper.tex
Final revision.

Most introductory resources, including~\cite{rojas_tutorial_2015}, present Lambda Calculus with named bindings.
Kiselyov's algorithms~\cite{kiselyov_lambda_2018} use De Bruijn indices~\cite{de_bruijn_lambda_1972} instead.
As this is core to the algorithm, we adopt the same approach.
However, for example expressions, we continue to use the easier-to-read named bindings.

@@

Combinatory Logic is a notation designed to remove variables (named or otherwise) from Lambda Calculus expressions.
The core idea is to define a set of simple basic functions, or \emph{combinators}, that can be combined to pass variables to the proper subexpressions.
We only need three combinators to convert any Lambda Calculus expression:
\begin{itemize}
    \item $S \ f \ g \ x = f \ x \ (g \ x)$: Passes an argument to both subexpressions of an application.
    \item $K \ x \ y = x$: Drops the second argument.
    \item $I \ x = x$\footnote{$SKK = I$, so even this combinator is not strictly necessary, but all reasonable implementations do include $I$.}: The identity function that accepts an argument and returns it.
\end{itemize}

@@

\begin{equation*}
    \begin{array}{l l}
        C\ [[\ \lambda x.\lambda y. x\ ]]        & \Rightarrow \\
        A\ x\ [[\ C\ [[\ \lambda y. x\ ]]\ ]]    & \Rightarrow \\
        A\ x\ [[\ A\ y\ [[\ C\ [[\ x\ ]]\ ]]\ ]] & \Rightarrow \\
        A\ x\ [[\ A\ y\ [[\ x\ ]]\ ]]            & \Rightarrow \\
        A\ x\ [[\ K\ x\ ]]                       & \Rightarrow \\
        S\ (A\ x\ [[\ K\ ]])\ (A\ x\ [[\ x\ ]])  & \Rightarrow \\
        S\ (K\ K)\ I                             & \Rightarrow \\
    \end{array}
\end{equation*}

@@

At this point, it could be that the graph has been reduced to a constant value (e.g. 9).
Another possibility is parts of the graph are still reduceable, but our reduction strategy does not `reach' it.
For example, we can have the following graph (including a $K$ combinator) that our strategy will not reduce:

@@

Benchmarks were run on an otherwise quiescent laptop equipped with an x86-64 AMD CPU (Ryzen 7 4800U), running Debian 11 (Bullseye).
The original Miranda was compiled using GCC 10.2.1 without any optimization flags, to avoid breaking the garbage collector as discussed in Section~\ref{sec:miranda}.
All other code is written in Rust and compiled using the \texttt{release} profile (\texttt{rustc} version 1.65.0).

@@

\section{Conclusion}
% Repeat the main question / objective of your research
Our objective with this work was to revive research around Combinator Graph Reduction by improving the performance.
We used a novel semantic translation algorithm to compile lambda calculus programs into smaller combinator expressions~\cite{kiselyov_lambda_2018} and have implemented modern combinator graph reduction engines based on the designs of Miranda~\cite{turner_new_1979} and TIGRE~\cite{koopman_fresh_1989}.

>>>>

Most introductory resources, including~\cite{rojas_tutorial_2015}, present Lambda Calculus with named bindings.
Kiselyov's algorithms~\cite{kiselyov_lambda_2018} use De Bruijn indices~\cite{de_bruijn_lambda_1972} instead.
As this is core to the algorithm, we adopt the same approach for our algorithm descriptions in Section~\ref{sec:kiselyov}.
However, in the example expressions we give in this paper, we continue to use the easier-to-read named bindings.

@@

Combinatory Logic is a notation designed to remove variables (named or otherwise) from Lambda Calculus expressions.
The core idea is to define a set of basic functions, or \emph{combinators}, that can be combined to pass variables to the proper subexpressions.
We only need three combinators to convert any Lambda Calculus expression:
\begin{itemize}
    \item $S \ f \ g \ x = f \ x \ (g \ x)$: Passes an argument to both subexpressions of an application.
    \item $K \ x \ y = x$: Drops the second argument.
    \item $I \ x = x$\footnote{$SKK = I$, so even this combinator is not strictly necessary, but all reasonable implementations do include $I$.}: The identity function that accepts an argument and returns it.
\end{itemize}

@@

\begin{equation*}
    \begin{array}{l l}
        C\ [[\ \lambda x.\lambda y. x\ ]]        & \Rightarrow \\
        A\ x\ [[\ C\ [[\ \lambda y. x\ ]]\ ]]    & \Rightarrow \\
        A\ x\ [[\ A\ y\ [[\ C\ [[\ x\ ]]\ ]]\ ]] & \Rightarrow \\
        A\ x\ [[\ A\ y\ [[\ x\ ]]\ ]]            & \Rightarrow \\
        A\ x\ [[\ K\ x\ ]]                       & \Rightarrow \\
        S\ (A\ x\ [[\ K\ ]])\ (A\ x\ [[\ x\ ]])  & \Rightarrow \\
        S\ (K\ K)\ I                             &             \\
    \end{array}
\end{equation*}

@@

At this point, it could be that the graph has been reduced to a constant value (e.g. 9).
Another possibility is that parts of the graph are still reduceable, but our reduction strategy does not `reach' it.
For example, we can have the following graph (including a $K$ combinator) that our strategy will not reduce:

@@

Benchmarks were run on an otherwise quiescent laptop equipped with an x86-64 AMD CPU (Ryzen 7 4800U), running Debian 11 (Bullseye).
The original Miranda was compiled using GCC 10.2.1 without any optimization flags, to avoid breaking the garbage collector as discussed in Section~\ref{sec:miranda}.
All other code is written in Rust and compiled using the optimizing \texttt{release} profile (\texttt{rustc} version 1.65.0).

@@

\section{Conclusion}
% Repeat the main question / objective of your research
Our objective with this work was to revive research around Combinator Graph Reduction by improving the performance.
We used a recently proposed semantic translation algorithm to compile lambda calculus programs into smaller combinator expressions~\cite{kiselyov_lambda_2018} and have implemented modern combinator graph reduction engines based on the designs of Miranda~\cite{turner_new_1979} and TIGRE~\cite{koopman_fresh_1989}.
